---
title: "Choose Your Own Project: Pancreatic ductal adenocarcinoma biomarkers"
author: "Daniel Flores Orozco"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4, scipen = 999)
```

# 1. Introduction 

Pancreatic ductal adenocarcinoma (PDAC) is the most common pancreatic neoplasm and is considered a highly aggressive and lethal cancer (Sarantis et al., 2020). In fact, less than 20% of individuals diagnosed with PDAC survive the first year, and less than 10% survived more than five years (Debernardi et al., 2020; Sarantis et al., 2020). The poor prognosis is mainly attributed to late diagnosis and the lack of efficient treatments during an early stage. However, when the disease is detected in the early stages, the probability of surviving more than five years increases up to 70% (Debernardi et al., 2020). Therefore, the development of novel tests for the early detection of PDAC is essential to improve the patients' prognosis. 

Debernardi et al. (2020) identified a group of proteins (creatinine, LYVE1, REG1A, REG1B, TFF1, and plasma CA19-9) in the urine that could potentially serve as biomarkers to detect PDAC. Debernardi et al. (2020) developed an algorithm to differentiate cancer samples from benign samples using LYVE1, REG1B, and TFF1 levels, as well as creatinine levels and age as predictors. Using logistic regression, they achieved a sensitivity and specificity of 0.963 and 0.967, respectively.

In this document, different Machine Learning algorithms, including random forest, kNN, SVM, logistic regression, and bayesian logistic regression, were used to predict if a given sample belonged to a benign or cancer group using the data provided by Debernardi et al. (2020). The ultimate objective of this project was to improve the performance reported by the original authors.  



# 2. Methodology
```{r load data, echo=FALSE, message=FALSE, warning=FALSE}
load("rda/raw_data.rda")

```


## 2.1 Database
The pancreatic cancer biomarkers data was downloaded from the database repository Kaggle (https://www.kaggle.com/johnjdavisiv/urinary-biomarkers-for-pancreatic-cancer). The original data contain 14 columns: *`r colnames(raw_data)`*. A description of each column information is presented in **Table 1**. The database contained 590 entries of the same number of patients. The patients' ages range from 26-89 years, with 299 males and 291 females. Around 33.7% of the entries came from patients diagnosed with cancer, and the rest from healthy or other benign pancreatic diagnostic.   

```{r columns details, echo=FALSE, message=FALSE, warning=FALSE}
#load file 
load("rda/columns_details.rda")

#load kable library 
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")
library(kableExtra)

#print
column_details %>% 
  kbl(caption = "Columns details in original data set", booktabs = TRUE) %>%
  kable_classic(latex_options = "hold_position") %>% 
  kable_styling(full_width = FALSE) %>%
  column_spec(1, italic = TRUE) %>% 
  column_spec(2, width = "30em")

```

Since the information *patient_cohort*, *sample_origin*, was not useful for the development of the different models, they were removed. Similarly, *stage*, and *bening_sample_diagnosis* were not included as the number of samples in each group was small. 


## 2.2 Data exploration 
The first step was to explore how the different predictors were distributed in the diagnosis groups (healthy, benign, and cancer) and determine if all of them could be good predictors. 

### 2.2.1 Diagnosis by sex 
The first approach consisted of exploring PDAC diagnosis was similar in male and female patients. A contingency table combined with a Chi-squared test were used to evaluate this. The results indicated that cancer diagnosis was significantly (p-value = 0.0001) more common in males than females. For instance, 58.3% of the cancer group were males.

### 2.2.2 Diagnosis by age
The distribution of diagnosis by age was also evaluated (Fig.1). This analysis revealed that cancer was most common in older people. Samples from people between 65-85 were more likely to be from the cancer group. For instance, more than 93% of cancer cases came from people older than 50 years. 

```{r fig1, echo=FALSE, warning=FALSE, echo=FALSE}
#load fig.1 
load("figs/fig.1")
fig.1

```


### 2.2.3 Biomarkers by diagnosis group and sex 

The evaluation of the concentration of the levels of the different biomarkers revealed that all of them, except creatinine, were overexpressed in the cancer groups compared to healthy and benign groups, as seen in **Fig. 2**. Moreover, **Fig.2** also shows that there is no noticeable difference in levels of the different biomarkers between males and females.

```{r fig2, echo=FALSE, warning=FALSE, message=FALSE}
#load
load("figs/fig.2")
#print
fig.2

```

### 2.2.4 Biomarkers by age and diagnosis group
Figure 3 shows the average concentration of each of the biomarkers in cancer, benign, and healthy groups at different ages. This figure shows that the creatine levels are similar and steady across all ages, which suggests that this protein may not be a good predictor. On the other hand, the levels of the other biomarkers seem to be constantly higher in cancer samples across all ages. These differences are substantially larger in older patients, which appears to result from a drop of these proteins in healthy patients rather than an increase in sick patients. 

```{r biomarkers by ageanddiag, echo=FALSE, warning=FALSE, fig.height=10, fig.width=8}
#load
load("figs/fig.3")
#print
fig.3

```

### 2.2.5 Samples' distance 
The final approach consisted in ordinating the samples with multidimensional scaling or principal coordinate analysis (PCoA) using the Bray-Curtis distance (Fig 4). For this, samples were classified as cancer or benign to simplify the analysis. Fig. 4a shows that when considering the biomarkers only, the two PC account for 59.1% of the variability (PC1 = 36.9%; PC2 = 22.2%). When including the patients' age (Fig.4b), the variability explained by the two PC increased to 70.4% (PC1-50.2%; PC2-22.2%), suggesting that age could be a good predictor. Although part of the variability of the data could be described for all these predictors, Fig.4 shows an overlapping of a significant fraction of cancer and benign samples. Fig.4 also shows there is no evident difference between male and female samples in the groups.  

```{r fig4, echo=FALSE, warning=FALSE, message=FALSE}
#load package 
if(!require(ggpubr)) install.packages("ggpubr", repos = "http://cran.us.r-project.org")
library(ggpubr)

#load figs 
load("figs/fig.4a")
load("figs/fig.4b")

#arrange Fig.4
ggarrange(fig.4a, fig.4b, 
          labels = c("a", "b"), 
          ncol = 1, nrow = 2) %>%
  annotate_figure("Figure 4. PCoA using Bray-Curtis distance")


```


## 2.3 Loss funtion 
The performance of the models was evaluated based on the proportion of cases correctly identified as patient with cancer or benign condition (overall accuracy) as well as the sensitivity and specificity of the model, which were defined as follows: 

$$
sensitivity: TPR = \frac{TP}{TP+FN}
$$
$$
specificity: TNR = \frac{TN}{TN+FP}
$$
Where $TPR$ and $TNR$ represent the true positive rate and true negative rate, respectively; $TP$ represents the true positive results, $FN$ false negative results, $TN$ true negative results, and $FP$ false positive results. 

## 2.4 Models 
The available data was fit to several different models including the  general linear model (GLM), bayesian GLM, k-nearest neighbors (KNN), linear discriminat analysis (LDA), quadratic discriminat analysis (QDA), random forest (rf), and a support-vector machine (svm) model. In a first approach, only the 4 biomarkers present in all samples (LYVE1, REG1B, TFF1, and creatinine) plus patient's age were used as predictors. In a second approach, all biomarkers and age were used as predictors, thus, only samples with information on all biomarkers were included (N = 209). 
Since data exploration revealed that suggested that the sex of the patient was not a strong predictor,it was not included in the model developed in this project. 

### 2.4.1 Models with 4 biomarkers 
For these models, the entire data set was split into train and test set. The train set was composed of 70% of the data whereas the test set of the other 30%. All models were developed using the training set only, and then evaluated with the test set.

### 2.4.2 Models with 6 biomarkers 
For these models, the subset of samples with reported concentration of all six biomarkers (N=209) were also randomly divided into training (70%) and test sets (30%). 


# 3. Results 

The performance of the models using age, LYVE1, REG1B, and TFF1 as predictors is shown on Table 3. 

```{r table3, echo=FALSE, echo=FALSE, warning=FALSE}
load("rda/model_perf.rda")

model_perf %>% 
  kbl(caption = "Performance of model with three biomarkers and age", booktabs = TRUE) %>%
  kable_classic()

```



# 4. Conclusion 


# 5. References  
